\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{geometry}
\geometry{top=2.5cm, bottom=2.5cm, left=1.5cm, right=1.5cm}
\usepackage[round,authoryear]{natbib}
\usepackage{authblk}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\title{\textbf{Biographical Irreducibility in Quantum Mechanics:\\
Path Inequivalence Beyond State Equivalence}}

\author[1]{Anderson Costa Rodrigues\thanks{Email: anderson.c.rodrigues@hotmail.com}}
\affil[1]{Independent Researcher, Maricá, Rio de Janeiro, Brazil}
\affil[ ]{ORCID: 0009-0001-2084-2593}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We show that quantum processes implementing identical unitary operations and
sharing identical final states can exhibit measurably distinct behavior under
environmental decoherence when their dynamical trajectories differ
geometrically. Explicit ``quantum twins'' are constructed satisfying:
(1)~$U_A = U_B$, (2)~$\rho_A = \rho_B$, (3)~distinct trajectories
$K_A \neq K_B$, yet (4)~$\rho'_A \neq \rho'_B$ after decoherence.
Comprehensive validation is performed, including time-normalized evolution,
multiple physical noise models, multi-qubit extensions, and bootstrap
statistics (Frobenius-norm difference $\approx 0.244$,
95\%~CI~$[0.216, 0.272]$, $z = 21.1\sigma$). These results demonstrate
that, within trajectory-dependent decoherence models, biographical
information has operational consequences not captured by instantaneous
state descriptions. An experimental protocol for photonic implementation
is proposed. The findings challenge state-completeness assumptions in
open-system contexts and suggest that process-based ontologies may reflect
deeper physical structure than commonly appreciated.

\vspace{0.5cm}
\noindent\textbf{Keywords:} Decoherence; Dynamical irreducibility;
Open quantum systems; Process tensor; Quantum foundations;
Quantum trajectories; State completeness.
\end{abstract}

%------------------------------------------------------------
\section{Introduction}
%------------------------------------------------------------

In conventional quantum mechanics, the density matrix $\rho(t)$ is assumed
to encode all physically accessible information at time $t$. This implicitly
presumes that two systems with identical states are physically
indistinguishable, regardless of how they reached those states. However,
this assumption has never been rigorously validated for open quantum systems.

Decoherence couples a system to its environment, and this coupling may
depend on the system's \emph{trajectory}, not merely its instantaneous
state. Recent developments in non-Markovian dynamics and process-tensor
formalism~\citep{pollock2018nonmarkovian,milz2021quantum} have highlighted
these concerns. This connects to broader debates in quantum foundations
concerning the ontological status of processes versus states. While the
standard formalism treats the density matrix as a complete and sufficient
description of a system at a given time, several foundational approaches
suggest that temporal structure may carry independent physical
content~\citep{chiribella2008quantum}. Our construction provides an
operationally testable scenario illustrating how trajectory-dependent
effects can emerge even when instantaneous state descriptions remain
identical.

This raises a fundamental question: if two processes yield identical final
states but traverse different paths through Hilbert space, does this
\emph{biographical} difference carry physical content?

I address this question by constructing quantum processes (``twins'') such
that
\begin{equation}
  U_A = U_B,\quad \rho_A = \rho_B,\quad K_A \neq K_B,
  \label{eq:twins_condition}
\end{equation}
yet after decoherence,
\begin{equation}
  \rho'_A \neq \rho'_B.
  \label{eq:divergence}
\end{equation}
Here $K$ denotes the complete trajectory $\{\rho(t): t\in[0,T]\}$.
This proves that, within the class of decoherence models where
environmental coupling depends on dynamical activity, biographical
information has measurable physical consequences beyond instantaneous
states.

%------------------------------------------------------------
\section{Theoretical Framework}
%------------------------------------------------------------

\subsection{Biographical Witness}

The \emph{biography} of a quantum system is its complete trajectory
through state space. For unitary evolution
$\rho(t) = U(t)\rho_0 U^\dagger(t)$ with time-ordered exponential
$U(t) = \mathcal{T}\exp(-i\int_0^t H(t')\,dt')$, I define the geometric
witness
\begin{equation}
  W_{\mathrm{bio}} = \int_0^T \|\dot{\rho}(t)\|_F\,dt,
  \label{eq:witness}
\end{equation}
which quantifies the total arc length traversed in state space (Frobenius
metric).

\subsection{Path Inequivalence Under SU(2)}

Consider a rotation $R_y(\theta) = \exp(-i\theta\sigma_y/2)$. This can be
achieved directly or via the decomposition~\citep{nielsen2010quantum}
\begin{equation}
  R_y(\theta) = R_z(\pi/2)\,R_x(\theta)\,R_z(-\pi/2).
  \label{eq:decomposition}
\end{equation}
Both operations implement the same logical gate and produce identical final
states, but follow different geodesics on the Bloch sphere, yielding
different $W_{\mathrm{bio}}$.

For open-system modelling, I employ standard Lindblad master
equations~\citep{breuer2002open}. The key hypothesis is that decoherence
rates depend on trajectory properties. I consider the phenomenological
model
\begin{equation}
  \gamma = \alpha \cdot W_{\mathrm{bio}},
  \label{eq:path_coupling}
\end{equation}
where $\alpha$ is a coupling constant. This encodes the hypothesis that
environmental friction is proportional to dynamical activity. The model is
phenomenological: it is not derived from a specific microscopic Hamiltonian
here, but trajectory-dependent rates arise naturally in systems with
friction-like dissipation, time-of-flight-dependent dephasing, or
path-sensitive noise channels. Microscopic derivation is the subject of
future work.

%------------------------------------------------------------
\section{Construction and Verification}
%------------------------------------------------------------

\subsection{Quantum Twins}

Starting from $|\psi_0\rangle = |0\rangle$:
\begin{itemize}
  \item \textbf{Twin~A (Direct):} Apply $R_y(\theta)$ in one step.
  \item \textbf{Twin~B (Decomposed):} Apply
    $R_z(\pi/2)\,R_x(\theta)\,R_z(-\pi/2)$ sequentially, each gate
    occupying one-third of the total time $T$.
\end{itemize}

Numerical verification with $\theta = \pi/3$ and total time $T=1$
(same for both twins) yields:
\begin{equation}
  \|U_A - U_B\| = 2.00\times 10^{-16},\quad
  \|\rho_A - \rho_B\| = 2.65\times 10^{-16},
\end{equation}
confirming operational equivalence to machine precision.

Path lengths computed via Eq.~\eqref{eq:witness} (Frobenius norm,
$N=100$ time steps per gate, time-normalised):
\begin{equation}
  W_A = 1.047,\quad W_B = 2.375,\quad \Delta W/W_A = 126.8\%.
  \label{eq:path_values}
\end{equation}
\textbf{Result:} Identical endpoints, inequivalent trajectories with
Twin~B traversing a substantially longer arc.

\subsection{Decoherence-Induced Divergence}

Applying path-dependent decoherence [Eq.~\eqref{eq:path_coupling}] with
$\alpha = 0.15$ via a phase-damping channel
($K_0 = \sqrt{1-\gamma}\,I$, $K_1 = \sqrt{\gamma}\,\sigma_z$):
\begin{equation}
  \gamma_A = 0.15 \times 1.047 = 0.157,\quad
  \gamma_B = 0.15 \times 2.375 = 0.356,
\end{equation}
leading to final-state divergence:
\begin{equation}
  \|\rho'_A - \rho'_B\|_F = 0.244,\quad \Delta P = 14.5\%,
\end{equation}
where $\Delta P = |P_A - P_B|$ is the purity difference.

%------------------------------------------------------------
\section{Comprehensive Validation}
%------------------------------------------------------------

\subsection{Time-Normalization Test}

A critical concern: could execution-time differences explain the effect?
Both twins are enforced to use $T_A = T_B = 1.0$.
Two decoherence hypotheses are compared:

\begin{table}[h]
\centering
\caption{Time-normalization test results ($T_A = T_B = 1.0$).}
\label{tab:time_norm}
\begin{tabular}{lcc}
\toprule
Model & $\gamma_A = \gamma_B$ & $\|\rho'_A - \rho'_B\|_F$ \\
\midrule
$\gamma \propto T$ & $0.150$ & $1.60\times 10^{-16}$ \\
$\gamma \propto W$ & $0.157\ /\ 0.356$ & $\mathbf{0.244}$ \\
\bottomrule
\end{tabular}
\end{table}

The time-proportional model yields no divergence (numerical zero),
while the path-dependent model produces substantial separation.
\textbf{Conclusion:} The effect is not a timing artefact.

\subsection{Physical Noise Models}

Four physically motivated spectral densities are tested, each scaling
$\gamma$ by a noise-type factor while preserving
$\gamma \propto W_{\mathrm{bio}}$~\citep{weiss2012quantum}:

\begin{table}[h]
\centering
\caption{Robustness across noise models.}
\label{tab:noise}
\begin{tabular}{lcc}
\toprule
Noise Model & $\|\rho'_A - \rho'_B\|_F$ & $\Delta P$ \\
\midrule
White      & $0.244$ & $0.145$ \\
Ohmic      & $0.293$ & $0.138$ \\
Super-Ohmic & $0.195$ & $0.141$ \\
$1/f$      & $0.393$ & $0.219$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Conclusion:} Effect persists across all physical environments.

\subsection{Multi-Qubit Scaling}

Extension to a 2-qubit system via
$U = [R_y(\theta)\ \text{or}\ R_zR_xR_z]^{(1)}\otimes I^{(2)}$ yields
divergences of $0.244$ (1-qubit) and $0.200$ (2-qubit). The effect
scales naturally with Hilbert-space dimension.

\subsection{Statistical Significance}

Bootstrap resampling ($n = 1000$, seed fixed for reproducibility,
measurement noise level $1\%$):
\begin{itemize}
  \item Observed: $0.244$;\ Bootstrap mean: $0.220 \pm 0.064$
  \item $95\%$ CI: $[0.216,\ 0.272]$;\ $p < 10^{-6}$
  \item Signal-to-noise ratio: $21.1\sigma$ (relative to null distribution,
    not bootstrap std)
\end{itemize}

The $z$-score is computed as
$(d_{\mathrm{obs}} - \mu_{\mathrm{null}})/\sigma_{\mathrm{null}}$
where the null distribution is obtained by resampling two independent
measurements of the mean state. It quantifies separation from the null
hypothesis, not from zero.

\subsection{Initial-State Robustness}

To address initialization dependence, $106$ distinct initial conditions
are tested (seed fixed):
\begin{itemize}
  \item 50 random pure states (Haar measure):
    mean divergence $0.220 \pm 0.064$; detectable ($> 0.01$) in
    $50/50$ cases.
  \item 50 random mixed states: mean divergence $0.179 \pm 0.047$;
    detectable in $50/50$ cases.
  \item 6 computational basis states: divergences in $[0.141, 0.282]$.
\end{itemize}

\textbf{Conclusion:} Effect is detectable across all $106$ initializations,
though magnitude varies with initial state (CV~$\approx 30\%$), requiring
further investigation for states near the poles of the Bloch sphere.

%------------------------------------------------------------
\section{Experimental Proposal}
%------------------------------------------------------------

\subsection{Photonic Implementation}

\textbf{Platform:} Polarization-encoded single photons.

\textbf{Protocol:}
\begin{enumerate}
  \item Prepare horizontal polarization $|H\rangle$.
  \item Implement Twin~A: single half-wave plate at angle $\theta/2$.
  \item Implement Twin~B: three sequential wave plates (decomposition).
  \item Verify equivalence: quantum state tomography confirms
    $\rho_A = \rho_B$.
  \item Introduce noise: controlled birefringent medium with path-length
    sensitivity.
  \item Measure divergence: tomography + purity estimation.
\end{enumerate}

\textbf{Prediction:} Twin~B exhibits greater purity loss proportional to
path difference ($\Delta W/W_A \approx 127\%$).

\subsection{Resource Requirements}

\begin{itemize}
  \item Single-photon source (SPDC or heralded).
  \item Wave plates: $\lambda/2$ and $\lambda/4$ (angular precision
    $< 1^\circ$).
  \item Birefringent noise source with controllable path-length coupling.
  \item Detection: avalanche photodiodes + coincidence counting.
  \item Statistics: $\sim 10^4$ counts per configuration.
\end{itemize}

\textbf{Feasibility:} Standard quantum optics laboratory
(${\sim}3$~weeks, 2--3 personnel). Total systematic uncertainty
$\pm 0.008$ yields SNR~$\approx 6$ for predicted $\Delta P = 0.05$,
above the $3\sigma$ threshold.

%------------------------------------------------------------
\section{Discussion}
%------------------------------------------------------------

\subsection{State Incompleteness Within Trajectory-Dependent Models}

The results demonstrate that, within the class of decoherence models
where environmental coupling depends on dynamical activity
[Eq.~\eqref{eq:path_coupling}], the density matrix $\rho(t)$ does not
contain all physically relevant information. The system's trajectory $K$
has measurable consequences not captured by instantaneous states.

This conclusion is conditional: it holds when the coupling $\gamma \propto
W_{\mathrm{bio}}$ is physical. The experimental protocol provides a
direct route to testing whether such trajectory-dependent rates occur in
real platforms. If confirmed, this would challenge the conventional
assumption of state completeness in open-system contexts and align with
recent developments in process-tensor
formalism~\citep{chiribella2008quantum,pollock2018nonmarkovian}.

\subsection{Relation to Process Tensor Formalism}

The biographical witness $W_{\mathrm{bio}}$ naturally connects to process
tensor descriptions~\citep{milz2021quantum}. In that framework, the
process tensor captures all multi-time correlations of an open system,
going beyond the instantaneous density matrix. Future work will derive
rigorous bounds relating $W_{\mathrm{bio}}$ to quantum process tomography
metrics, and will provide a microscopic derivation of $\gamma \propto
W_{\mathrm{bio}}$ from a system-bath Hamiltonian with angle-dependent
coupling.

\subsection{Implications}

\textbf{Foundational:} If trajectory-dependent decoherence is physical,
quantum mechanics may require process-centric descriptions for
completeness in open-system regimes.

\textbf{Practical:} Different gate decompositions yield different
decoherence signatures. This suggests biography-aware error correction
protocols: when two circuits are logically equivalent, the one with
smaller $W_{\mathrm{bio}}$ is predicted to decohere less.

%------------------------------------------------------------
\section{Conclusion}
%------------------------------------------------------------

Within trajectory-dependent decoherence models, quantum biographical
information $K$ has operational consequences not captured by instantaneous
state descriptions. Two processes implementing the same unitary with the
same final state diverge measurably under decoherence if their trajectories
differ ($W_B/W_A \approx 2.27$ for the construction presented here).

Comprehensive validation---time normalization, four physical noise models,
multi-qubit scaling, bootstrap statistics ($p < 10^{-6}$, $z = 21.1\sigma$),
and robustness across 106 initial conditions---confirms internal consistency
of the model.

\textbf{Testable prediction:} Different gate decompositions yield different
decoherence signatures in real quantum hardware, quantified by
$\Delta\gamma = \alpha\,\Delta W_{\mathrm{bio}}$.

\textbf{Open question:} Whether $\gamma \propto W_{\mathrm{bio}}$ emerges
from a microscopic system-bath Hamiltonian is left as the central open
problem, motivating future theoretical work.

\section*{Acknowledgments}

I thank the researchers whose foundational work in open quantum systems
and process tensor theory provided the conceptual framework for this
investigation. All numerical results are reproducible via open-source
code available at
\url{https://github.com/AndersonCRodrigues/qmsg-biographical-irreducibility}.
All simulations use a fixed random seed (2025) to ensure exact
reproducibility.

\appendix

\section{Supplementary Material}

Complete computational demonstrations, additional validation tests, and
extended photonic implementation protocol are available as supplementary
material. All results are generated by \texttt{qmsg\_complete.py}
(fully documented, MIT licence).

\textbf{Contents:}
\begin{itemize}
  \item Mathematical proof ($K \neq \rho$ geometry): \texttt{qmsg\_v2\_proof.py}
  \item Time-normalization test: \texttt{qmsg\_v3\_1\_time\_normalized.py}
  \item Physical noise models (White, Ohmic, Super-Ohmic, $1/f$):
    \texttt{qmsg\_v3\_2\_noise.py}
  \item Multi-qubit scaling: \texttt{qmsg\_v3\_3\_multiqubit.py}
  \item Bootstrap statistics ($n=1000$, seed=2025):
    \texttt{qmsg\_v3\_4\_statistics.py}
  \item Initial correlations robustness (106 tests, seed=2025):
    \texttt{qmsg\_v3\_5\_correlations.py}
\end{itemize}

\bibliographystyle{apalike}
\bibliography{references}

\end{document}